How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985

### 2022-09-14 Draft_September_2022/Full_draft_20220901.rmd
#correctionsagain

*We need to protect the Amazon from foreign interests. We need to exploit the Amazon's natural resources. We need to provide better living standards for the people in the Amazon. We need to preserve the Amazon as a standing ecosystem.* Each of these statements contains an implicit assumption of what needs to be solved, or in other words, it represents the Amazon as a particular problem: national sovereignty, economic integration, social development, and environmental conservation, respectively.These specific problems touch on common understandings of the Amazon as part of the larger socio-cultural history of the country.

Different governments throughout Brazilian history have been described as proponents of a specific view of the Amazon [@drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. We conceptualize the Amazon region, forest, and people to be a policy object, this is a specific issue that deserves dedicated policy attention. While the military dictatorship(1964-85) is associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney's (1985-1990) and Lula's (2003-2010)presidencies are often tied to environmental conservation views. These monolithic representations advance the view that specific governments understand and act towards the Amazon as one problem, hiding within government diversity of both policies and perspectives. Albeit the current calls to understand the environment as a social-cultural construction and to identify the effect of culture on environmental outcomes [@lepolaindewaroux2021], we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments. Building on @hirschman1963 conceptualization of chosen problems and @bacchi2009 theory of problem-representation in policy, we propose a framework for identifying problem-constructions in discourse and investigate how the Brazilian Amazon has been constructed as a problem in transnational political discourses. We understand transnationalism[^1] as relationships that transcend nation-states, from the local level to the supranational level encompassing non-state actors [@keck1998].

@@

Although problem-construction takes place in a series of instances, we analyze the case of speeches by Brazilian presidents since 1985. Scholarly research about environmental policy in Brazil has focused on the role of the environmental bureaucracy [@silva-muller2022a], non-governmental organizations [@keck1998], legislation [@soares-filho2014], and international markets [@assuncao2015], and many others. Nevertheless, the role of the president remains understudied. We opt for presidential speeches for two reasons. First, presidential discourses have the power to introduce and justify the public policy, as well as shape its perception to broad audiences [@zarefsky2004; @gillion2016]. It legitimizes ways of thinking about the Amazon. In turn, policy perception is key for policy adoption and implementation [@alesina2009; @lopez2020]. This is especially pertinent to deforestation in Brazil as expectations of local actors, generated from material and discursive governmental practices, are a crucial factor in decisions to deforest on the ground [@assuncao2015; @capobianco2019; @capobianco2021, @campbell2015]. Second, we argue that presidents can deploy problem-constructions that build policy objects as different problems depending on whom they are speaking to. Presidential discourses take place in different settings, from launching a new bridge in a small municipality in the middle of the Amazon to a speech at the UN general assembly in New York. Working with presidential discourses allows us to identify this variation in meaningful ways and better understand how the Amazon is discursively constructed in transnational politics.

@@

Our empirical findings are threefold. First, we find that the frequency of the Amazon as a topic in speeches inside and outside Brazil is equally driven by domestic events (e.g. the assassination of Chico Mendes) as well as international events (e.g. Climate Summits). Second, the same government adopts a diverse range of problem-constructions over time. While economic integration dominated discourse from 1985 to the mid-2000s, environmental conservation and social development constructions steadily grew in the 2000s, briefly surpassing economic integration as more common problem-constructions from 2011 to 2016. Constructing the Amazon as an issue of sovereignty has increasingly become more pertinent after 2010. Finally, using logit models, we find that presidents are more likely to construct the Amazon as a problem of environmental conservation as presidents move farther away from the Amazon region. That is, presidents talk to the economic and social needs of the people domestically while boasting about environmental policy internationally. Conceptually, this contributes to understanding the social construction of the Amazon in discourses and how these relates to policies. Empirically, we provide the first comprehensive overview of when, where, and how the Amazon is constructed as problems in presidential speeches.

@@

In "Journey towards Progress", @hirschman1963 draws a conceptual distinction between pressing problems (pressured by outside parties to the government) and chosen problems (chosen by the government at their discretion). Pressing problems can be either privileged or neglected, depending on the degree of pressure exercised by the interested groups. Chosen problems are those that governments select at their discretion. Problems can change from pressing to chosen as a function of (a) solutions becoming available, (b) a change in the level of government control in society, or (c) a shift of interests from top policymakers [@hirschman1963]. @hirschman1975 's example of policymakers choosing a problem is President's Kubitschek decision of building Brasilia., which was not particularly pressed by any interest group.

@@

To analyze our data, we first present a series of different plots on the proportions of Amazonian statements and problem-constructions over time and by presidents. To test whether problem-constructions depend on where presidents speak, we run four separate logistic regressions. Each model takes the share of a specific problem-construction versus all other problem-constructions as their dependent variable. This means we try to predict the probability of a president constructing the Amazon as an issue of (1) environmental conservation, (2) economic integration, (3) social development, and (4) national sovereignty, separately, and in relation to the remaining three problem-constructions. Our main independent variable is the distance (in kilometers) from  the state or country capital where the speech took place to Manaus (the capital of the Amazonas State and centrally located in the region). In our model, we also control for annual deforestation rates, annual inflation rates, and election years in the model. We interpret the plots and model considering multiple Amazon-related events and policies over the last 30 years, with the support of existing literature, and their correlations with different presidents and locations.

@@

From 2010 to 2018, except for 2015, we see a general decrease in mentions of the Amazon in all official presidential speeches. From 2010 to 2014, mentions of the Amazon went down from 18% to 8%. Disagreements related to the priority of environmental preservation over economic development, most notably in the case of approval of the New Forest Code in 2012, were frequent. The 2012 Forest Code regularized land ownership of many illegally deforested areas. The period was also marked by political and economic instability which culminated in the impeachment of Rousseff in 2016. In 2015, the Amazon appeared in 19% of all speeches. This maximum coincides with the Paris summit which became a key turn in climate politics after the failures of Copenhagen. Brazil went to the Paris Summit with deforestation numbers slightly higher than Copenhagen and a perception that there was a turn towards less conservation inside the country. By 2016 mentions of Amazon in official speeches went down to 6%, the lowest share since 1985.

@@

The trends identified are especially interesting considering Amazonian policies and deforestation rates portrayed in Figure 3, above. First, deforestation rates have a strong negative correlation with environmental conservation. When deforestation rates increased from 1988 to 2004, and from 2012 until 2021, environmental conservation constructions decreased. As deforestation rates decreased from 2004 to 2012, environmental conservation constructions in presidential speeches increased. @capobianco2021 argues that the unprecedented decrease in deforestation we observed from 2004 to 2012 was a product of an increase in the perception of stronger federal policies and police presence in the Amazon region which, in turn, generated a perception of a higher risk of being caught and fined for deforestation. The shift from constructing the Amazon as an issue of economic integration to an issue of environmental conservation generates a perception that illegal deforestation will be increasingly monitored and sanctioned. Second and relatedly, deforestation rates have a positive and strong correlation with economic integration. When deforestation increased from 1988 to 2004 and from 2012 to 2021, so did economic integration. When deforestation decreased from 2004 to 2012, economic integration decreased as well. This suggests that presidents might boast policy outcomes related to economic integration as justifications for deforestation.

@@

Table 1, below, shows a statistically significant relationship between constructing the Amazon as an issue of environmental conservation and distance from the Amazon. That is, the farthest away the president is from the Amazon region, the more likely the president is to construct it as an issue of environmental conservation. Second, the opposite is true for both economic integration and social development, with statistically significant negative relationships. When speaking within the Amazon, presidents focus on social development and economic integration. This entails that conservation might be a desirable construction when speaking internationally about the Amazon, but not at the local level, where presidents highlight economic and social development. This multi-level game implies that presidents talk to the economic and social needs of the people domestically while boasting about environmentalism internationally.

>>>>

*We need to protect the Amazon from foreign interests. We need to exploit the Amazon's natural resources. We need to provide better living standards for the people in the Amazon. We need to preserve the Amazon as a standing ecosystem.* Each of these statements contains an implicit assumption of what needs to be solved, or in other words, it represents the Amazon as a particular problem: national sovereignty, economic integration, social development, and environmental conservation, respectively.These specific problems touch on common understandings of the Amazon as part of the larger socio-cultural history of the country.Different governments throughout Brazilian history have been described as proponents of a specific view of the Amazon [@drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021].

We conceptualize the Amazon region, forest, and people to be a policy object, this is a specific issue that deserves dedicated policy attention. While the military dictatorship(1964-85) is associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney's (1985-1990) and Lula's (2003-2010)presidencies are often tied to environmental conservation views. These monolithic representations advance the view that specific governments understand and act towards the Amazon as one problem, hiding within government diversity of both policies and perspectives. Albeit the current calls to understand the environment as a social-cultural construction and to identify the effect of culture on environmental outcomes [@lepolaindewaroux2021], we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments. Building on @hirschman1963 conceptualization of chosen problems and @bacchi2009 theory of problem-representation in policy, we propose a framework for identifying problem-constructions in discourse and investigate how the Brazilian Amazon has been constructed as a problem in transnational political discourses. We understand transnationalism[^1] as relationships that transcend nation-states, from the local level to the supranational level encompassing non-state actors [@keck1998].

@@

Although problem-construction takes place in a series of instances, we analyze the case of speeches by Brazilian presidents since 1985. Scholarly research about environmental policy in Brazil has focused on the role of the environmental bureaucracy [@silva-muller2022a], non-governmental organizations [@keck1998], legislation [@soares-filho2014], and international markets [@assuncao2015], and many others. Nevertheless, the role of the president remains understudied. We opt for presidential speeches for two reasons. First, presidential discourses have the power to introduce and justify the public policy, as well as shape its perception to broad audiences [@zarefsky2004; @gillion2016]. It legitimizes ways of thinking about the Amazon. In turn, policy perception is key for policy adoption and implementation [@alesina2009; @lopez2020]. This is especially pertinent to deforestation in Brazil as expectations of local actors, generated from material and discursive governmental practices, are a crucial factor in decisions to deforest on the ground [@assuncao2015; @capobianco2019; @capobianco2021; @campbell2015]. Second, we argue that presidents can deploy problem-constructions that build policy objects as different problems depending on whom they are speaking to. Presidential discourses take place in different settings, from launching a new bridge in a small municipality in the middle of the Amazon to a speech at the UN general assembly in New York. Working with presidential discourses allows us to identify this variation in meaningful ways and better understand how the Amazon is discursively constructed in transnational politics.

@@

Our empirical findings are threefold. First, we find that the frequency of the Amazon as a topic in speeches inside and outside Brazil is equally driven by domestic events (e.g. the assassination of Chico Mendes) as well as international events (e.g.Climate Summits). Second, the same government adopts a diverse range of problem-constructions over time. While economic integration dominated discourse from 1985 to the mid-2000s, environmental conservation and social development constructions steadily grew in the 2000s, briefly surpassing economic integration as more common problem-constructions from 2011 to 2016. Constructing the Amazon as an issue of sovereignty has increasingly become more pertinent after 2010. Finally, using logit models, we find that presidents are more likely to construct the Amazon as a problem of environmental conservation as presidents move farther away from the Amazon region. That is, presidents talk to the economic and social needs of the people domestically while boasting about environmental policy internationally. Conceptually, this contributes to understanding the social construction of the Amazon in discourses and how these relates to policies. Empirically, we provide the first comprehensive overview of when, where, and how the Amazon is constructed as problems in presidential speeches.

@@

In "Journey towards Progress", @hirschman1963 draws a conceptual distinction between pressing problems (pressured by outside parties to the government) and chosen problems (chosen by the government at their discretion). Pressing problems can be either privileged or neglected, depending on the degree of pressure exercised by the interested groups. Chosen problems are those that governments select at their discretion. Problems can change from pressing to chosen as a function of (a) solutions becoming available, (b) a change in the level of government control in society, or (c) a shift of interests from top policymakers [@hirschman1963]. @hirschman1975 example of policymakers choosing a problem is President's Kubitschek decision of building Brasilia, which was not particularly pressed by any interest group.

@@

To analyze our data, we first present a series of different plots on the proportions of Amazonian statements and problem-constructions over time and by presidents. To test whether problem-constructions depend on where presidents speak, we run four separate logistic regressions. Each model takes the share of a specific problem-construction versus all other problem-constructions as their dependent variable. This means we try to predict the probability of a president constructing the Amazon as an issue of (1) environmental conservation, (2) economic integration, (3) social development, and (4) national sovereignty, separately, and in relation to the remaining three problem-constructions. 
Our main independent variable is the distance (in kilometers) from the state or country capital where the speech took place to Manaus (the capital of the Amazonas State and centrally located in the region). We interpret the plots and model considering multiple Amazon-related events and policies over the last 30 years, with the support of existing literature, and their correlations with different presidents and locations.

@@

From 2010 to 2018, except for 2015, we see a general decrease in mentions of the Amazon in all official presidential speeches. From 2010 to 2014, mentions of the Amazon went down from 18% to 8%. Disagreements related to the priority of environmental preservation over economic development were frequent, most notably in the case of approval of the New Forest Code in 2012, which regularized land ownership of many illegally deforested areas. The period was also marked by political and economic instability which culminated in the impeachment of Rousseff in 2016. In 2015, the Amazon appeared in 19% of all speeches. This maximum coincides with the Paris summit which became a key turn in climate politics after the failures of Copenhagen. Brazil went to the Paris Summit with deforestation numbers slightly higher than Copenhagen and a perception that there was a turn towards less conservation inside the country. By 2016 mentions of Amazon in official speeches went down to 6%, the lowest share since 1985.

@@

The trends identified are especially interesting considering Amazonian policies and deforestation rates portrayed in Figure 3, above. First, deforestation rates have a strong negative correlation with environmental conservation. When deforestation rates increased from 1988 to 2004, and from 2012 until 2021, environmental conservation constructions decreased. As deforestation rates decreased from 2004 to 2012, environmental conservation constructions in presidential speeches increased. @capobianco2021 argues that the unprecedented decrease in deforestation we observed from 2004 to 2012 was a product of an increase in the perception of stronger federal policies and police presence in the Amazon region which, in turn, engendered a perception of a higher risk of being caught and fined for deforestation. The shift from constructing the Amazon as an issue of economic integration to an issue of environmental conservation generates a perception that illegal deforestation will be increasingly monitored and sanctioned. Second and relatedly, deforestation rates have a positive and strong correlation with economic integration. When deforestation increased from 1988 to 2004 and from 2012 to 2021, so did economic integration. When deforestation decreased from 2004 to 2012, economic integration decreased as well. This suggests that presidents might boast policy outcomes related to economic integration as justifications for deforestation.

@@

Table 1, below, shows a statistically significant relationship between constructing the Amazon as an issue of environmental conservation and distance from the Amazon. That is, the farthest away the president is from the Amazon region, the more likely the president is to construct it as an issue of environmental conservation. Second,the opposite is true for both economic integration and social development, with statistically significant negative relationships. When speaking within the Amazon, presidents focus on social development and economic integration. This entails that conservation might be a desirable construction when speaking internationally about the Amazon, but not at the local level, where presidents highlight economic and social development. This multi-level game implies that presidents talk to the economic and social needs of the people domestically while boasting about environmentalism internationally.
